## Writer verification with a Siamese network
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,       # Show code by default
  message = FALSE,   # Suppress package startup messages globally
  warning = FALSE,   # Suppress warnings globally
  fig.align = "center" 
)
```


<!-- Task 3: contrastive or triplet loss; how training examples are constructed; matching performance. -->

A Siamese network was trained on the handwriting images to evaluate how accurately it could capture whether two digits were written by the same writer. 
To do so, we had to control for the digit identity by constructing pairs using the same digit. Therefore, a positive pair for this network consists of the same digit done by the same writer (but different images), and a negative pair is the same digit but a different writer. Examples of pairs from the training and validation datasets is shown in @fig-pairs.

We conducted balanced sampling for the pairs, so that half of the pairs came from the same person and half different (check; rephrase).

Contrastive loss was used - (do I need a justification?) - which introduces another hyperparameter for model tuning: the margin, $m$. The formula is:
$$
L = yD^2 + (1-y)\max(0,m-D)^2
$$
where $y = 1$ for a positive pair (same writer, same digit) and $D$ is the Euclidean distance between the embeddings. $m$ sets the distance at which embeddings for similar pairs results in zero loss (where the distance reaches the margin) versus embeddings for dissimilar pairs resulting in a large loss, thereby optimising the network to minimise differences between image pairs from the same writer and maximising differences between image pairs from different writers @ELARABY20221837. The margin for this initial Siamese network was set as 1 [following the notes], but was optimised in (ref section 5).

The training/validation/test data splitting was done on writer-level to ensure that the network learns the writing style, or writer identity, and so that the generalisability of the model can be tested (accurately/with minimal leakage) with writers it hasn't seen in training or validation. However, since there are only 20 writers, a 60/20/20 split leaves only 4 writers for validation and testing independently, which is a significant limitation and is expanded in (ref section 4).

The encoder architecture was set as a simple CNN with two convolutional layers, max pooling and global pooling...
[because the input images are low-resolution and contain (easily learnable) features like curves and strokes] 

The Siamese model outputs distances between the two embeddings for each digit pair, which is then turned into a classification of "same" or "different" writer using a threshold determined by the validation pairs [also a limitation with the few writers & the split we have]. We compared the balanced accuracy for actual same and different digit pairs at different thresholds, and chose the threshold with the maximised balanced accuracy. 

The generalisability of our model was evaluated by predicting on unseen writers' digit image-pairs. The results are shown in @tab-confusionmat and @tab-testresults. @fig-test-results shows the test batch's predicted and true results


```{r readdata}

handwriting <- readRDS("../data/handwriting.rds")
#head(handwriting$metadata)

library(keras3)
use_backend("tensorflow")

# images is an integer array with values from 0 (background) to 255 (ink).

# below = x in e.g.
x <- handwriting$images / 255 # scale the pixels to 0-1
#dim(handwriting$images) # 25000 x 28 x 28 x 1 - 1 = number of image channels (greyscale = 1)
# therefore: 25,000 images, 28 x 28 pixels, 1 image channel
#range(x) # 0 1

# below = 'labels' in e.g.
y_person <- handwriting$metadata$person_id - 1L # subtracts 1 from every value in ID
#range(y_person) # 0 19 = 20 people

set.seed(25)
```


```{r makepairs}

# need pairs of image indices to compare
# use person_id to decide if positive / negative pair

make_pairs <- function(ids, # indices of images it can use
                       n, # how many pairs to create
                       metadata){ # metadata -> writer ID + digit ID for each image
  
  # a: choose the first image in every pair - randomly selects n IDs
  a <- sample(ids, n, replace = TRUE) # w/replacement = can make as many pairs as we want
  
  # b: create the second image in every pair
  b <- vapply(a, function(i){
    
    # randomly decide if pair is from same writer or not
    same <- runif(1) < 0.5 
    
    # finds pool of images with the same digit
    pool <- ids[metadata$digit[ids] == metadata$digit[i]]
    
    if(same){ # if the same writer, pool must be same writer + same digit (pos pair)
      
      # 'from all images of digit i, keep only the ones written by the same writer'
      # CHANGE TO Y_PERSON OR NOT?
      pool <- pool[metadata$person_id[pool] == metadata$person_id[i]]# who wrote image i
      
    } else { # if not the same writer, pool must be diff writer + same digit (neg pair)
      
      pool <- pool[metadata$person_id[pool] != metadata$person_id[i]]
      
    }
    sample(setdiff(pool, i), 1) # doesn't allow pairing with itself
    
  }, integer(1)) # each iteration must return 1 integer
  
  cbind(a,b)
}

```


```{r makebatch}

make_batch <- function(p, # pairs created by make_pairs()
                       images = x, # actual images
                       class_labels = y_person) { # class assoc with images
  
  y <- matrix(as.integer( # true = 1, false = 0
    class_labels[p[, 1]] # labels of first col of p-i matrix
                         == class_labels[p[, 2]]), # label of second col
              ncol = 1) # matrix of 1 col for formatting for neural network
  
  # return the actual images + the target
  list(
    left = images[p[, 1], , , , drop = FALSE], # left side of Siamese
    right = images[p[, 2], , , , drop = FALSE], # right side of Siamese
    y = y
  )
}

```



```{r datasplit}
# get the writer IDs belonging in each split
set.seed(25)
writers <- unique(handwriting$metadata$person_id - 1L)
train_writers <- sample(writers, 0.6*length(writers))

rest_writers <- setdiff(writers, train_writers)
set.seed(25)
valid_writers <- sample(rest_writers, 0.5*length(rest_writers))
test_writers <- setdiff(rest_writers, valid_writers)

# now extract the image IDs belonging to each split
train_ids <- which(handwriting$metadata$person_id %in% train_writers)
valid_ids <- which(handwriting$metadata$person_id %in% valid_writers)
test_ids <- which(handwriting$metadata$person_id %in% test_writers)

# now make the image index pairs in each split
train_pairs <- make_pairs(train_ids, length(train_ids), handwriting$metadata)
valid_pairs <- make_pairs(valid_ids, length(valid_ids), handwriting$metadata)
test_pairs <- make_pairs(test_ids, length(test_ids), handwriting$metadata)

# for each pair, extract the image + 0/1
train <- make_batch(train_pairs)
valid <- make_batch(valid_pairs)
test  <- make_batch(test_pairs)
```

Images of pairs: 

```{r}
#| label: fig-pairs
#| fig-cap: "Example training and validation pairs. Top row: training pairs (same writer left, different writer right). Bottom row: validation pairs (same writer left, different writer right)."
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(2, 4), mar = c(1, 1, 2, 1))

show_pair <- function(pair_idx, pairs, label) {
  img1 <- x[pairs[pair_idx, 1], , , 1]
  img2 <- x[pairs[pair_idx, 2], , , 1]
  image(t(img1[nrow(img1):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = label)
  image(t(img2[nrow(img2):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = "")
}

pos_train <- which(train$y[, 1] == 1)
neg_train <- which(train$y[, 1] == 0)
pos_valid <- which(valid$y[, 1] == 1)
neg_valid <- which(valid$y[, 1] == 0)

show_pair(pos_train[1], train_pairs, "Train: same writer")
show_pair(neg_train[1], train_pairs, "Train: different writer")
show_pair(pos_valid[1], valid_pairs, "Val: same writer")
show_pair(neg_valid[1], valid_pairs, "Val: different writer")
```


```{r encoderbuild}

image <- layer_input(shape = c(28, 28, 1)) # each image = 28x28, 1 image channel

z <- image |>
  layer_conv_2d(16, 3, activation = "relu") |> # learns useful features (shapes, strokes)
  layer_max_pooling_2d(2) |> # makes feature maps smaller while keeping important info
  layer_conv_2d(32, 3, activation = "relu") |>
  layer_global_average_pooling_2d() |> # converts feature maps -> single vector of numbers
  layer_dense(16) # converts to 16-no embedding representing the handwriting style

encoder <- keras_model(image, z)
# NOTHING HAS BEEN CHANGED FOR THE ABOVE FROM THE NOTES -- CHANGE IF WANTED
```

```{r siamesebuild}
# a & b = describes pair of images supplied to model
a <- layer_input(shape = c(28, 28, 1))
b <- layer_input(shape = c(28, 28, 1))

# encoders produce the embeddings for a & b
# distance calculates the Euclidean distance between their embeddings 
distance <- list(encoder(a), encoder(b)) |>
  layer_lambda(f = function(z) {
    op_sqrt(op_sum(op_square(z[[1]] - z[[2]]), 
                   axis = 2, # sum over embedding dimension
                   keepdims = TRUE) + 
              1e-7) # makes sure it's not 0
  })

# distance = model output
model <- keras_model(list(a, b), distance)
```


```{r constrastiveloss}

contrastive_loss <- function(y, d) {
  op_mean( # finds avg loss across all pairs in a batch
    y * op_square(d) + # squares each pair's distance - for same-writer pairs, penalises a large distance
            (1 - y) * op_square(op_maximum(1 - d, 0))) # for diff writer pairs, penalises only if distance is < 1: if dist >=1, penalty is 0
}

```

```{r compile}
model |> compile(
  optimizer = optimizer_adam(1e-3),
  loss = contrastive_loss
)

model |> fit(
  list(train$left, train$right), train$y,
  validation_data = list(
    list(valid$left, valid$right), valid$y
  ),
  epochs = 5,
  batch_size = 64,
  verbose=0
)
```

```{r valid_dist}

valid_distance <- as.numeric(predict(model, list(valid$left, valid$right), verbose = 0))
valid_same <- as.logical(valid$y[, 1])

balanced_accuracy_at <- function(cut) {
  predicts_same <- valid_distance < cut
  same_rate <- mean(predicts_same[valid_same])
  different_rate <- mean(!predicts_same[!valid_same])
  (same_rate + different_rate) / 2
}

#example_thresholds <- unique(as.numeric(
#  quantile(valid_distance, c(.25, .5, .75))
#))
#data.frame(
#  threshold = example_thresholds,
#  balanced_accuracy = vapply(example_thresholds, balanced_accuracy_at, numeric(1))
#)

threshold_candidates <- sort(unique(valid_distance))
threshold_scores <- vapply(threshold_candidates, balanced_accuracy_at, numeric(1))
best_index <- which.max(threshold_scores)
threshold <- threshold_candidates[best_index]

data.frame(threshold = threshold, balanced_accuracy = threshold_scores[best_index])
```


Test: -- change below !!
```{r}
#| label: tab-confusionmat

test_distance <- as.numeric(predict(
  model, list(test$left, test$right), verbose =0))

true_same <- as.logical(test$y[, 1]) # real answer if same/diff writer
predicted_same <- test_distance < threshold # uses model distance & best threshold 


# confusion matrix
confusion_matrix <- table(
  Actual = factor(true_same,
                  levels = c(FALSE, TRUE),
                  labels = c("Different writer", "Same writer")),
  Predicted = factor(predicted_same,
                     levels = c(FALSE, TRUE),
                     labels = c("Different writer", "Same writer"))
)

confusion_matrix
```


```{r}
#| label: tab-testresults


# extract values from confusion matrix
TN <- confusion_matrix["Different writer", "Different writer"]
FP <- confusion_matrix["Different writer", "Same writer"]
FN <- confusion_matrix["Same writer", "Different writer"]
TP <- confusion_matrix["Same writer", "Same writer"]

# Performance metrics
accuracy <- (TP + TN) / sum(confusion_matrix)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
specificity <- TN / (TN + FP)

results <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "Specificity"),
  Value = c(accuracy, precision, recall, specificity)
)

results$Value <- sprintf("%.1f%%", results$Value * 100)

results
```

```{r}
#| label: fig-test-results
#| fig-cap: "Sample test pairs with true label, predicted label, and distance score. Green border = correct prediction, red = incorrect."
#| fig-width: 10
#| fig-height: 6

set.seed(42)
n_show <- 8
sample_idx <- sample(seq_len(nrow(test_pairs)), n_show)

par(mfrow = c(2, n_show), mar = c(0.5, 0.5, 2.5, 0.5))

for (i in sample_idx) {
  img1 <- x[test_pairs[i, 1], , , 1]
  img2 <- x[test_pairs[i, 2], , , 1]
  
  true_label   <- ifelse(test$y[i, 1] == 1, "Same", "Different")
  pred_label   <- ifelse(test_distance[i] < threshold, "Same", "Different")
  correct      <- true_label == pred_label
  border_col   <- ifelse(correct, "forestgreen", "red3")
  score        <- round(test_distance[i], 3)
  
  title_text <- paste0(
    "True: ", true_label, ",", " Pred: ", pred_label, "\nd=", score
  )
  
  # left image
  image(t(img1[nrow(img1):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = title_text,
        cex.main = 0.75)
  box(col = border_col, lwd = 2)
  
  # right image
  image(t(img2[nrow(img2):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = "")
  box(col = border_col, lwd = 2)
}
```

