## Writer verification with a Siamese network

<!-- Task 3: contrastive or triplet loss; how training examples are constructed; matching performance. -->

A Siamese neural network was trained on the handwriting images and evaluated to assess how accurately it could capture whether two digits were written by the same writer. We controlled for digit identity by constructing pairs consisting of the same digits. Therefore, a positive pair for this network consisted of two different images of the same digit written by the same person, and a negative pair contained two images of the same digit but by different writers. Examples of pairs from the training and validation datasets are shown in @fig-pairs. We employed a balanced pair sampling approach with half of the pairs being positive and the remaining half negative to ensure the Siamese network learned effectively [@chopra_2005].

```{r}
#| label: fig-pairs
#| fig-cap: "Example training and validation pairs. Top row: training pairs (same writer left, different writer right). Bottom row: validation pairs (same writer left, different writer right)."

par(mfrow = c(2, 4), mar = c(1, 1, 2, 1))

show_pair <- function(pair_idx, pairs, label) {
  img1 <- x[pairs[pair_idx, 1], , , 1]
  img2 <- x[pairs[pair_idx, 2], , , 1]
  image(t(img1[nrow(img1):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = label)
  image(t(img2[nrow(img2):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = "")
}

pos_train <- which(train$y[, 1] == 1)
neg_train <- which(train$y[, 1] == 0)
pos_valid <- which(valid$y[, 1] == 1)
neg_valid <- which(valid$y[, 1] == 0)

show_pair(pos_train[1], train_pairs, "Train: same writer")
show_pair(neg_train[1], train_pairs, "Train: different writer")
show_pair(pos_valid[1], valid_pairs, "Val: same writer")
show_pair(neg_valid[1], valid_pairs, "Val: different writer")
```

Here we used contrastive loss, a loss function introduced by @hadsell_2006, in the Siamese network. This loss function is more appropriate compared to triplet loss since the distinction between similar and dissimilar pairs is of interest rather than the relative ranking of dissimilarity of triplets. Additionally, this is a relatively simple verification task not a complex classification task like facial recognition, and the dataset is relatively small, thus the model should converge more easily using contrastive loss [@herman_2026]. The contrastive loss for a pair is:

$$
L = yD^2 + (1-y)\max(0,m-D)^2
$$
where $y$ is a binary label with $y = 1$ for a positive pair (i.e. same writer, same digit) and $D$ is the Euclidean distance between the embeddings. The hyperparameter $m$ sets the distance at which embeddings for similar pairs results in zero loss (i.e. where the distance reaches the margin) versus embeddings for dissimilar pairs resulting in a large loss, thereby optimising the network to minimise differences between image pairs from the same writer and maximising differences between image pairs from different writers [@Elaraby_2023]. The margin for the initial Siamese network was set as one, but was optimised in a hyperparameter search in [Hyperparameter optimisation](#hyperparameter-optimisation). 

The data were split into 60% training, 20% validation, and 20% test sets. The data were grouped by writer with different writers in each set to ensure that the network learnt to recognise similarities in writing style but without leakage of specific writers to the validation and test set to test generalisability of the model. However, since there were only 20 writers, this split meant only 4 writers in each of the validation and test sets, which is a significant limitation and expanded on in [Train, validation and test strategy](#train-validation-and-test-strategy). 

The shared CNN encoder architecture consisted of two convolutional layers with 16 and 32 filters respectively, both with a filter size of 3 and a ReLu activation function to introduce non-linearity, a max pooling layer between these convolutional layers with both a pool size and stride of 2, and at the end, a global average pooling layer and a dense layer. The convolutional layers learn local image features while max pooling reduces the dimensions of the feature maps. Global pooling converts the feature maps into a vector which is mapped to a 16-number embedding by the dense layer. A relatively simple encoder is used since the input images are low-resolution and contain easily learnable features such as curves and strokes [ref].

In the Siamese model, pairs of images are supplied to the encoder to produce their embeddings and then the Euclidean distance between each pair is calculated. The model is compiled using the Adam optimiser and using the above contrastive loss function which is calculated using the Euclidean distances and a margin of one, and averaged for every batch. Epoch is set to 5 and batch size to 64. 

The outputs from the Siamese model are the Euclidean distances between the two embeddings for each digit pair, which is then turned into the classification of "same" or "different" writer using a threshold determined by the validation pairs [also a limitation with the few writers & the split we have]. We compared the balanced accuracy for actual same and different digit pairs from the validation set at different thresholds, and chose the threshold which maximised the balanced accuracy. The different thresholds tested were all the actual distances between the validation pairs. 

The generalisability of our model was evaluated by predicting on unseen writers' digit image-pairs in the test set. The results are shown in @fig-confusionmat and @tbl-testresults. Examples of pairs from the test batch with predicted and true values are shown in @fig-test-results 

```{r}
#| label: encoderbuild

set_random_seed(25) # set random seed for tensorflow

# Build shared encoder - the same as the notes to keep it simple 
image <- layer_input(shape = c(28, 28, 1)) # each image = 28x28, 1 image channel

z <- image |>
  layer_conv_2d(16, 3, activation = "relu") |> # learns useful features (shapes, strokes)
  layer_max_pooling_2d(2) |> # makes feature maps smaller while keeping important info
  layer_conv_2d(32, 3, activation = "relu") |>
  layer_global_average_pooling_2d() |> # converts feature maps -> single vector of numbers
  layer_dense(16) # converts to 16-number embedding representing the handwriting style

encoder <- keras_model(image, z)

```

```{r}
#| label: siamesebuild

# a & b = describes pair of images supplied to model
a <- layer_input(shape = c(28, 28, 1))
b <- layer_input(shape = c(28, 28, 1))

# encoders produce the embeddings for a & b
# distance calculates the Euclidean distance between their embeddings 
distance <- list(encoder(a), encoder(b)) |>
  layer_lambda(f = function(z) {
    op_sqrt(op_sum(op_square(z[[1]] - z[[2]]), 
                   axis = 2, # sum over embedding dimension
                   keepdims = TRUE) + 
              1e-7) # makes sure it's not 0
  })

# distance = model output
model <- keras_model(list(a, b), distance) 
```

```{r}
#| label: compile

# Use Adam optimiser to compile model and contrastive loss function defined earlier
model |> compile(
  optimizer = optimizer_adam(1e-3),
  loss = contrastive_loss # using fn from siameseFunctions.R
)

# Fit model with training data and validation data
model |> fit(
  list(train$left, train$right), train$y,
  validation_data = list(
    list(valid$left, valid$right), valid$y
  ),
  epochs = 5,
  batch_size = 64,
  verbose=0
)
```

```{r}
#| label: valid_dist

# Calculate the model predicted distances between the pairs in the validation data
valid_distance <- as.numeric(predict(model, list(valid$left, valid$right), verbose = 0))
# Extract the observed labels for whether the pairs are the same or different for validation data
valid_same <- as.logical(valid$y[, 1])

# Function to calculate balanced accuracy at different thresholds
balanced_accuracy_at <- function(cut) {
  predicts_same <- valid_distance < cut # at given threshold (cut) would model predict same or different?
  same_rate <- mean(predicts_same[valid_same]) # accuracy for "same" pairs
  different_rate <- mean(!predicts_same[!valid_same]) # accuracy for "different" pairs
  (same_rate + different_rate) / 2 # mean accuracy for "same" and "different" ("alanced accuracy)
}

threshold_candidates <- sort(unique(valid_distance)) # threshold candidates are based on the distances themselves
threshold_scores <- vapply(threshold_candidates, balanced_accuracy_at, numeric(1)) # calculate balanced accuracy for all threshold candidates using fn from siameseFunctions.R
threshold <- threshold_candidates[which.max(threshold_scores)] 
best_index <- which.max(threshold_scores) # index for the threshold that maximises the balanced accuracy
threshold <- threshold_candidates[best_index] # threshold that maximises the balanced accuracy

data.frame(threshold = threshold, balanced_accuracy = threshold_scores[best_index])
```

```{r}
#| label: fig-confusionmat
#| fig-cap: "Confusion matrix for predictions from the Siamese neural network on pairs of images in the test dataset."

# Calculate the model predicted distances between the pairs in the test data
test_distance <- as.numeric(predict(model, list(test$left, test$right), verbose =0))
# Extract the observed labels for whether the pairs are the same or different for test data
true_same <- as.logical(test$y[, 1]) 
# Predicted labels based on selected threshold
predicted_same <- test_distance < threshold # uses model distance & best threshold 

# Construct confusion matrix based on observed and predicted labels
confusion_matrix <- table(
  Actual = factor(true_same,
                  levels = c(TRUE, FALSE),
                  labels = c("Same writer", "Different writer")),
  Predicted = factor(predicted_same,
                     levels = c(TRUE, FALSE),
                     labels = c("Same writer", "Different writer"))
)

# Plot confusion matrix
confusion_matrix %>% as.data.frame() %>% # first convert to data frame
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) + # fill with counts
  geom_tile() +
  geom_text(aes(label = Freq), color = "black", size = 6) +
  scale_fill_gradient(low = "orange", high = "deepskyblue") + # add colour according to scale
  scale_y_discrete(limits = rev) + # swap categories on y axis
  scale_x_discrete(position = "top") + # move labels to top
  labs(x = "Predicted", y = "Actual") +
  theme_minimal() + 
  theme(panel.grid = element_blank(), legend.position = "none", # remove legend
        text = element_text(size = 16), # increase text size
        axis.text.y = element_text(angle = 90, hjust = 0.5)) # adjust y axis labels

```

```{r}
#| label: tbl-testresults
#| tbl-cap: "Assessment metrics for Siamese neural network predictions on pairs of images in the test set."

# Extract values from confusion matrix
TN <- confusion_matrix["Different writer", "Different writer"]
FP <- confusion_matrix["Different writer", "Same writer"]
FN <- confusion_matrix["Same writer", "Different writer"]
TP <- confusion_matrix["Same writer", "Same writer"]

# Performance metrics
accuracy <- (TP + TN) / sum(confusion_matrix)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
specificity <- TN / (TN + FP)

results <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "Specificity"),
  Value = c(accuracy, precision, recall, specificity)
)

results$Value <- sprintf("%.1f%%", results$Value * 100)

results %>% kable()
```

```{r}
#| label: fig-test-results
#| fig-cap: "Sample test pairs with true label, predicted label, and distance score. Green border = correct prediction, red = incorrect."

set.seed(38)
n_show <- 8
sample_idx <- sample(seq_len(nrow(test_pairs)), n_show)

par(mfrow = c(4, 4), mar = c(0.5, 0.5, 2.5, 0.5))

for (i in sample_idx) {
  img1 <- x[test_pairs[i, 1], , , 1]
  img2 <- x[test_pairs[i, 2], , , 1]
  
  true_label   <- ifelse(test$y[i, 1] == 1, "Same", "Different")
  pred_label   <- ifelse(test_distance[i] < threshold, "Same", "Different")
  correct      <- true_label == pred_label
  border_col   <- ifelse(correct, "forestgreen", "red3")
  score        <- round(test_distance[i], 3)
  
  title_text <- paste0("True: ", true_label, ",", " Pred: ", pred_label, "\nd = ", score)
  
  # left image
  image(t(img1[nrow(img1):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = title_text,
        cex.main = 1)
  box(col = border_col, lwd = 2)
  
  # right image
  image(t(img2[nrow(img2):1, ]), col = grey(seq(1, 0, length = 256)),
        axes = FALSE, main = "")
  box(col = border_col, lwd = 2)
}
```

